---
  title: "Bug decoder for R beginners"
  output: 
    html_document:
      toc: true
---
*Pssst:* R bug not here? Send the bug to Slack #rtvmworkshop, including ?rbug in your message so that we can search for it.  

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=TRUE, prompt=TRUE)
```
***
### <font color="red"> Error: unexpected ')' in "some.object <- 2)"</font>  
> **Explanation:** There is a parenthesis somewhere that is not matching. In the example above there is an extra closing parenthesis after the `2` without a sister open parenthesis.
```{r eval=FALSE}
some.object <- 2 # now this will work
```

***
### <font color="red"> Error in obj1 + obj2 : non-numeric argument to binary operator </font>  
> **Explanation:** Mathematical operations in R only work with numbers, is one of `obj1` or `obj2` a character?
```{r eval=FALSE}
obj1 <- 3
obj2 <- "hi"
obj1 + obj2 # bug!
obj2 <- "3" # watch out, 3 is between quotation marks so interpreted as a character
obj1 + obj2 # bug!
obj2 <- 3
obj1 + obj2 # no bug since both obj1 and obj2 are numbers now
```

***
### <font color="red"> Error: object 'yourvar' not found</font>  
>**Explanation:** The object 'yourvar' is not in the R environment. Did you assign it a value?
```{r eval=FALSE}
ls() # to view the objects currently in the R environment
```
```{r}
yourvar <- 2 # assign the value 2 to yourvar
yourvar # now print it
yourvar * 2 # success
```

If it is the column of a dataframe, make sure you specify that with `mytunadf$yourvar`.  

***  
### <font color="red"> Error in file(file, "rt") : cannot open the connection In addition: Warning message: In file(file, "rt") : cannot open file 'file.txt': No such file or directory</font>  
> **Explanation:** the file 'file.txt' is not in your current working directory.


```{r eval=FALSE}
# To see the current location of the working directory
getwd()
# To change the location of your working directory
setwd("/my/new/working/directory") # specify path here, e.g. "C:/Projects/OFP-data" (note the "")
# To see what files are in your current working directory
list.files() # is file.txt here?
```