---
title: "Data import"
output: 
  html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

# From Excel (.xls, .txt, .csv)

You can read in an external data set from a `.txt` or `.csv` file using the functions `read.table()` or `read.csv()`.
If your data is in Excel, the easiest option is first convert your `.xls` into a `.csv` (using Save As...). Once your spreadsheet has been converted to a `.csv`, you can import it and store it in a R dataframe as follows: 

```{r}
catch_effort <- read.csv(file="sample_data.csv") # your table is now stored in the object 'catch_effort'
head(catch_effort) # have a look at the first 6 rows
```


Ps: you can read in a `.xls` file directly but you will need to install the `gdata` library first, and you might have to install an additional software (Perl).

# From R (.RData)

If your data has already been formatted and saved as a R object, you can import it directly using `load()`. Files with the extension `.RData` contain R objects, load them with the argument `verbose=TRUE` so that the name of the objects gets printed out.

```{r}
load('catch.RData', verbose=TRUE)
```