---
title: "My first ggplot map"
output: 
  html_document: 
    toc: true

---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, highlight=TRUE, prompt=TRUE, cache=TRUE)
```
# Introduction
ggplot is a useful way to create maps and add layers to the map. 

In this example we will create a map, highlight the PNA Member EEZs and then overlay the regions of the South Pacific albacore assessment.

# Step1 

Load all the requited packages.

```{r}
library(survival)
library(maps)
library(maptools)
library(mapdata)
library(ggplot2)
library(plyr)
library(grid)
library(scatterpie)
```

# Step 2
Load all the mapping data you will need eg. the EEZs you want and the data to be plotted. 

```{r}
all_states <- map_data("world2Hires") # if you want to go above 180 use world2Hires else worldHires

#eez overlay
eez        <- read.table("EZNEW2.txt", sep=" ", header=F, skip=0) #All regions
eez_pna  <- read.csv("pna_eez.csv") #PNA regions

point_data<-read.csv("point_dat.csv") # the data for the points

```

# Step 3
Define the colours for the background and then load all the states (countries) you want to appear in the plot.

```{r}

#different made up colours for the sea
grecol    <- rgb(red=0,green=100, blue=0, alpha=40, maxColorValue=255) #green colour
blucol    <- rgb(red=60,green=200, blue=175, alpha=70, maxColorValue=255)

#sort(unique(all_states$region)) # to list the states you will need this if you want to add countries to your map

# All states for the map.
states     <- subset(all_states, region %in% 
                       c("American Samoa", "Australia", "Canada","China","Cook Islands","Fiji", "French Polynesia", 
                         "Guam", "Hawaii", "Indonesia","Japan","Kiribati", "Marshall Islands", "Mexico", "Micronesia",
                         "Nauru", "New Caledonia", "New Zealand", "South Korea","North Korea" , "Niue", 
                         "Northern Mariana Islands","Palau", 
                         "Papua New Guinea", "Philippines", "Samoa", "Solomon Islands","Tokelau", "Tonga", "Tuvalu", 
                         "USA", "USSR", "Vanuatu","Panama","Mongolia" , "Chile","Argentina", "New Caledonia","Belize",
                         "Nicaragua","Ecuador","Honduras","Costa Rica","Colombia", "Uruguay", "Brazil", 
                         "Peru", "Guatemala","Venezuela","Bolivia","Paraguay" ))
```

# Step 4
Load the co-ordinates for the regional boundaries and make the labels. 

```{r}
# add the region co-ordinates for the SP albacore assessment 
b.x_1 <- c(140,140,170,170,140)
b.y_1 <- c(-10,  0,  0,-10,-10)
b.x_2<- c(140,140,170,170,140)
b.y_2<- c(-25,-10,-10,-25,-25)
b.x_3<- c(140,140,170,170,140)
b.y_3<- c(-50,-25,-25,-50,-50)
b.x_4<- c(170,170,210,210,170)
b.y_4<- c(-10,  0,  0,-10, -10)
b.x_5<- c(170,170,210,210,170)
b.y_5<- c(-25,-10,-10,-25,-25)
b.x_6<- c(170,170,210,210,170)
b.y_6<- c(-50,-25,-25,-50,-50)
b.x_7<- c(210,210,230,230,210)
b.y_7<- c(-25, -5, -5,-25,-25)
b.x_8<- c(210,210,230,230,210)
b.y_8<- c(-50,-25,-25,-50,-50)

# region labels
reg_lab<-c("1","2","3","4","5","6","7","8") # labels
reg_labx<-c(155,155,155,190,190,190,220,220) # x axis co-ordinates
reg_laby<-c( -5,-18,-35, -5,-18,-35,-15,-35) # y axis co-ordinates
```

# Step 5
Now you are ready to start plotting. 
```{r}

# make the map this is compiled like a ggplot graph but it puts it together as a series of layers
p          <- ggplot() +
  # add in the polygons (the shape of the eezs)
  # note aes is the long (x) and lat (y) coordinates 
  # note add the eezs first then the countries over them
  geom_polygon(data=eez, aes(eez[,1], eez[,2]), colour="white", fill=grecol) + # All eezs
  geom_polygon(data=eez_pna, aes(eez_pna[,1], eez_pna[,2]), colour="white", fill="blue") + # PNA eezs
  geom_polygon(data=states, aes(x=long, y=lat, group = group),colour ="white",  fill="grey10" ) + # Add the countries
  
  # Add in the region boundaries
  geom_polygon(aes(x=b.x_1, y=b.y_1), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_2, y=b.y_2), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_3, y=b.y_3), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_4, y=b.y_4), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_5, y=b.y_5), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_6, y=b.y_6), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_7, y=b.y_7), color="grey60", size=1.5, fill=NA) +
  geom_polygon(aes(x=b.x_8, y=b.y_8), color="grey60", size=1.5, fill=NA) +
  
   # Add points (proportion from own region)
  geom_point(aes(x = point_data$lat, y = point_data$long), color="red", size=point_data$val, alpha=0.5)+
  
  # region labels
  annotate("text", label = reg_lab, x = reg_labx, y = reg_laby, size = 6, colour = "blue")+ # add in the text you want
  
  coord_cartesian(xlim =c(130,235),ylim=c(-55,40)) + #sets data limits on x and y
  theme(panel.background = element_rect(fill=blucol), 
        plot.title = element_text(size = 20),
        axis.text.x = element_text(vjust =1, size = 14, colour = "black"),
        axis.text.y = element_text(hjust =1, size = 14,  colour = "black"),
        axis.title.x = element_text(size=14),
        axis.title.y = element_text(size=14,angle=90)) +  
  xlab('long') + ylab('lat') 

p

#savePlot("test_map", type="png")

```

#Next we will plot some pie charts on a map 

As an example we will show the level of catch and the catch proportions averaged over the last 10 years for some PNA mambers longline fisheries

```{r}
library(survival)
library(maps)
library(maptools)
library(mapdata)
library(ggplot2)
library(plyr)
library(grid)
library(scatterpie)

all_states <- map_data("world2Hires") # if you want to go above 180 use world2Hires else worldHires

# To get the center of a country for plotting 
centres <- ddply(all_states,.(region),summarize,long=mean(long),lat=mean(lat))
head(centres)

#eez overlay
eez        <- read.table("EZNEW2.txt", sep=" ", header=F, skip=0) #All regions
eez_pna  <- read.csv("pna_eez.csv") #PNA regions

# Data for plotting 
pool_dat<-read.csv("pool_dat.csv", header = TRUE) # the data for the points

#different made up colours for the sea
grecol    <- rgb(red=0,green=100, blue=0, alpha=40, maxColorValue=255)    # green colour
blucol    <- rgb(red=60,green=200, blue=175, alpha=70, maxColorValue=255)

#sort(unique(all_states$region)) # to list the states, you will need this if you want to add countries to your map

# All states to be added to the map
states     <- subset(all_states, region %in% 
                       c("American Samoa", "Australia", "Canada","China","Cook Islands","Fiji", "French Polynesia", 
                         "Guam", "Hawaii", "Indonesia","Japan","Kiribati", "Marshall Islands", "Mexico", "Micronesia",
                         "Nauru", "New Caledonia", "New Zealand", "South Korea","North Korea" , "Niue", "Northern Mariana Islands","Palau", 
                         "Papua New Guinea", "Philippines", "Samoa", "Solomon Islands","Tokelau", "Tonga", "Tuvalu", 
                         "USA", "USSR", "Vanuatu","Panama","Mongolia" , "Chile","Argentina", "New Caledonia","Belize",
                         "Nicaragua","Ecuador","Honduras","Costa Rica","Colombia", "Uruguay", "Brazil", 
                         "Peru", "Guatemala","Venezuela","Bolivia","Paraguay" ))

# First make the map basic map and allocate it to the object called p
p  <- ggplot() +
  geom_polygon(data=eez, aes(eez[,1], eez[,2]), colour="white", fill=grecol) + # All eezs
  geom_polygon(data=eez_pna, aes(eez_pna[,1], eez_pna[,2]), colour="white", fill="blue") + # PNA eezs
  geom_polygon(data=states, aes(x=long, y=lat, group = group),colour ="white",  fill="grey10" ) + # Add the countries
  
  coord_cartesian(xlim =c(130,235),ylim=c(-55,40))+
  
  theme(panel.background = element_rect(fill=blucol), 
        plot.title = element_text(size = 20),
        axis.text.x = element_text(vjust =1, size = 14, colour = "black"),
        axis.text.y = element_text(hjust =1, size = 14,  colour = "black"),
        axis.title.x = element_text(size=14),
        axis.title.y = element_text(size=14,angle=90)) +  
  xlab('Long') + ylab('Lat')

```

You can check the map if you want to to make sure toyr base map is correct by typing p in the consol (as you have assigned the map to an object called "p")

# Next you will add the pie charts as layers to your map
Here we use some averaged data for 5 countries showing the average annual caych for the last 10 years and the species proportions of tuna in the longline catch. 
```{r}
# Add in the pie charts 
# if the pies are too big shrink the radius eg in this example I divide by 1500
windows(20,16) 
p + geom_scatterpie(aes(x=long, y=lat, group=region, r=tot/1500), data=pool_dat, cols=c("alb","bet","yft"),  color="black", alpha=.8) + # This adds on the pie chart  cols is the columns used in the pie chart.
    geom_scatterpie_legend(pool_dat$tot/1500, x=220, y=-50)   # this bit adds in the legend and places it at the x and y cordinates

#savePlot("pie_map", type="png")

```

# Changing the pie chart colours
Use the scale_fill_manual and add that in as a new layer
```{r}
windows(20,16) 
p + geom_scatterpie(aes(x=long, y=lat, group=region, r=tot/1500), data=pool_dat, cols=c("alb","bet","yft"),  color="black", alpha=.8) + 
  scale_fill_manual(values = c("forestgreen", "red", "yellow"))+ # To manually change the colours of the piechart
  geom_scatterpie_legend(pool_dat$tot/1500, x=220, y=-50)   # this bit adds in the legend and places it at the x and y cordinates

#savePlot("pie_map_cols", type="png")
```


