---
  title: "Bug decoder for R beginners"
  output:
  html_document:
   theme: sky
---
*Pssst:* R bug not here? Send the bug to Slack #rtvmworkshop, including ?rbug in your message so that we can search for it.  

***

> <font color="red"> Error: object 'yourvar' not found</font>  
**Explanation:** The object 'yourvar' is not in the R environment. Did you assign it a value?
```{r eval=FALSE}
ls() # to view the objects currently in the R environment
```
```{r}
yourvar <- 2 # assign the value 2 to yourvar
yourvar # now print it
yourvar * 2 # success
```

If it is the column of a dataframe, make sure you specify that with `mytunadf$yourvar`.  

***  

> <font color="red"> Error in file(file, "rt") : cannot open the connection In addition: Warning message: In file(file, "rt") : cannot open file 'file.txt': No such file or directory</font>  
**Explanation:** the file 'file.txt' is not in your current working directory.


```{r eval = FALSE}
# To see the current location of the working directory
getwd()
# To change the location of your working directory
setwd("/my/new/working/directory"") # specify path here, e.g. "C:/Projects/OFP-data" (note the "")
# To see what files are in your current working directory
list.files() # is file.txt here?
```