---
title: "Introduction to the ggplot2 package"
output:
  html_document:
    hightlight: tango
    toc: yes
    toc_float: true
---

```{r setup, include=FALSE}
# Global options
#
# echo = TRUE. Show the code
# comment = "". Don't put anything in front of results from code
# warning = FALSE. Don't display warning messages
# cache = TRUE. Only process new code when you knit
# dev = 'png'. Save graphics in figure folder as png
# dpi = 500. Resolution for saved figures
# fig.height = 3. Height of figures (in inches)

knitr::opts_chunk$set(echo = TRUE, 
                      comment = "",
                      warning = FALSE,
                      cache = TRUE,
                      dev = 'png', 
                      dpi = 500, 
                      fig.height = 3)
```

# Getting things ready

First we load the **ggplot2** package:
```{r start, message=FALSE}
# Load ggplot2 package and some others
library(ggplot2)
```

# Sample data set

Your working directory can be found with the command *getwd()*. This is the directory where data is loaded from and saved to by default. Commonly this directory is set to the same directory as where your R script or Markdown file is. 

If the data set is in your working directory it can be read in directly without giving the directory name. We assume the data set is in the working directory, read it in, and check the result with the `head` command. 

```{r read in data}
# catch and effort data
ce <- read.csv("sample.csv")
```

```{r look at data}
head(ce)
```

Ps: If the data set is in a different directory from your working directory then you need to tell the `read.csv` command where this directory is. 


# Inital plot setup

We wish to make a histogram and density plot for the column *ALB.mt* (monthly albacore tuna in metric tonnes). 

The initial setup for these plots is done with the **ggplot** function, which by itself it produces an empty plot. 

```{r ggplot setup}
ggplot(data = ce, aes(x = ALB.mt))  
```

The purpose of the *ggplot* function is to

* state what data frame to use:  `data = ce`
* state what columns we wish to plot:  `aes(x = ALB.mt)`

The argument `aes` is short for aesthetic and refers to general aspects of a plot (e.g. what is to be plotted, should the plot be broken up into groups such as months). You'll see more examples below to clarify what this means.

The code above says that we wish to plot the column `ALB.mt` from the data frame `ce`, but doesn't say how to do this plot (e.g. a histogram, density, scatter plot). To give a specific type of plot, further instructions are added to the code, as we will see in the following sections where we make histograms and density plots.  

# Histogram and density plots

In this section we make histograms and density plots for the albacore tuna catch.

A non-empty plot will be made once you state what sort of plot you wish to make (e.g. histogram or density plot), which is referred to as the *geometry* for the plot. In this case we wish to plot x (i.e. ALB.mt) as a histogram, so we use the function **geom_histogram**. Refer to the ggplot2 cheat sheet for the other type of plots you can make for a single column of data (called One Variable in the cheat sheet)

As choosing the bin width is often best done manually, to explore what patterns there are in the data, the function suggests using the *binwidth* argument to do this. The default is to use 30 bins for the plot.


```{r histogram}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_histogram()
```

You can give arguments to the geom_histogram function for the bin width and fill colour of the histogram. The default bin width looks pretty reasonable, but I've a chosen something different to illustrate setting the bin width. 

```{r plot histogram explicit}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_histogram(binwidth = 20, fill = "darkblue")
```

Now do a histogram with bins of size 10, and fill the rectangles with another colour.

Use `theme_set()` to change the default plot theme to have a white background.
```{r plot histogram explicit II}

theme_set(theme_bw()) # set all ggplots to have a black and white background

ggplot(data = ce, aes(x = ALB.mt)) +
  geom_histogram(binwidth = 10, fill = "tomato3")
```

In the next plot we explicitly define the bin widths using the *breaks* argument. The first bin goes from 0 to 20 and is centred on 10 metric tonnes.

To change the interval width in function `seq()`, modify the value of the `by=` argument to 50. 
```{r plot histogram explicit III}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_histogram(breaks = seq(from = 0, to = 350, by = 50), 
                 fill = "darkblue")
```

If we wish to make a density plot we change the "geometry" by using the **geom_density** function. The argument *adjust* for this function determines how smooth the density plot is, with higher values giving smoother plots. 


```{r plot density}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_density(adjust = 0.1, fill = "darkblue")  # play with the adjust argument to see what it does
```

```{r plot density II}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_dotplot(fill='darkblue', color='red')
```

# Scatterplots

## Introductory example

In this section we plot the big eye tuna catch against the albacore tuna catch. As before the initial plot setup is done with the **ggplot** function, adding in another argument for the big eye tuna catch. 


```{r ggplot intro setup, fig.show='hide'}
## Both x and y are continuous variables
## In the cheatsheet, look under the Two Variables/Continous X-Continuous Y panel
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  
```

The code above says that we wish to plot the columns "ALB.mt" and "BET.mt"" from the data frame "ce", but doesn't say how to do this plot (e.g. using points or lines).

As we wish to plot points we use the function **geom_point**. Refer to the ggplot2 cheat sheet to see what other types of plots can be made (under the heading **Continuous X, Continuous Y**). 


```{r ggplot scatterplot I}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  + geom_point()
```

We could also plot this using lines if we wanted to, using **geom_line** instead.
```{r}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  + geom_line()
```
(Ugh!)

We can even combine different plot types: 
```{r}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  +
  geom_line(color='grey') + geom_point(color='red') 
# can change individual features by layer
```

If we wanted to keep track of which months was which on the graph, we could specify in the `aes` function to also colour the points by month:
```{r}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt, color=Month))  +
  geom_point()
```

To treat the months as discrete values, add `as.factor(Month)` instead, this makes `R` treat the month numbers as categories:
```{r}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt, color=as.factor(Month)))  +
  geom_point() 
```

## Exercises

1. Make a histogram of yellowfin catch with appropriate bin widths. 
Bonus: Add properly labelled x and y axis.

```{r}
ggplot(data = ce, aes(x = YFT.mt)) +
  geom_histogram(binwidth = 5, fill = "gold") + 
  labs(x='Monthly yellowfin catch (mt)', y='Count', title='Yellowfin', subtitle='A very tasty tuna')
```

2. Make a density plot of bigeye catch, coloured with the bigeye WCPFC colour (`firebrick`)

```{r}
ggplot(data = ce, aes(x = BET.mt)) +
  geom_density(adjust = 0.5, fill = "firebrick", color=NA) + 
  labs(x='Monthly bigeye catch (mt)', y='Smoothed count', title='Bigeye tuna', 
       subtitle='A tuna previously thought to be overfished')
```

3. Make a scatter plot of yellowfin catch (Y) vs. bigeye catch (X). 
```{r}
ggplot(data = ce, aes(x = BET.mt, y = YFT.mt)) +
  geom_point() +
  ggtitle('Yellowfin vs. bigeye catch') 
```

4. Make a scatter plot of yellowfin catch by month, with the years coloured as categories.
```{r}
ggplot(data = ce, aes(x = Month, y = YFT.mt, color=as.factor(Yr))) +
  geom_point() +
  ggtitle("Monthly yellowfin catch") + ylab('Yellowfin catch (mt)')
```

**Bonus exercise (woop!)**

Plot the average yellowfin catch per month (use `dcast` to create a new dataset with the mean.) Don't call it Brad.

Step 1: Calculate the mean YFT catch per month (goal: get a table with a month column, and a column for mean YFT catch)
```{r}
library(reshape2)
## First read-in the dataset from this morning
ce.long <- read.csv('sample_data-long.csv')
## Now calculate the mean catch by species and month
tuna.mean <- dcast(ce.long, Month ~ Species, value.var='Catch.mt', fun.aggregate=mean)
head(tuna.mean)
```

Step 2: Make the plot with month as the x-variable and mean YFT as the y-variable
```{r}
ggplot(data=tuna.mean, aes(x=Month, y=ALB)) + geom_point() + geom_line()
```

#### Adding both YFT, BET and ALB on the same plot when we have individual columns for each

```{r}
ggplot(data=tuna.mean, aes(x=Month, y=ALB)) + geom_line(color='forestgreen') + 
  geom_line(aes(y=YFT), color='gold') + 
  geom_line(aes(y=BET), color='firebrick')
```

# Adding a trend line to the plot

Adding a smoothed line helps in following the pattern of the points, and this can be done with the function **geom_smooth**, with the default a *loess curve* with a 95% confidence interval. 

This is adding a linear regression
```{r ggplot scatterplot II}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  +
  geom_point() +
  geom_smooth(method='lm', colour = "red")
```

This is adding a smoother
```{r}
ggplot(data = ce, aes(x = ALB.mt, y = BET.mt))  +
  geom_point() +
  geom_smooth(colour = "red")
```

## Monthly albacore catch over time

The *ce* data frame has columns labelled year and month, but they are given as integers, and R does not know that 12 means December. We create a numeric year-month variable by transforming the months into fractions of the year.

```{r make date, results="hide" }
# Make a new column called date
ce$date <- ce$Yr + (ce$Month-1)/12
head(ce$date)
```

Now we the plot the catch over time. 

```{r plot albacore catch over time}
ggplot(ce, aes(x = date, y = ALB.mt)) + 
  geom_point() + 
  geom_line(colour = "brown", linetype = 2)

```

### Adding plot labels
You can easily add axes labels and a plot title with ggplot, using the commands `xlab`, `ylab` and `ggtitle`:

```{r}
ggplot(ce, aes(x = date, y = ALB.mt)) + 
  geom_point() + 
  geom_line(colour = "brown", linetype = 2) + 
  xlab("Date") + 
  ylab("Monthly albacore catch (metric tonnes)") +
  ggtitle("Albacore catch over time")

```



# Further examples and help on the internet

Solutions to common graphing problems

[http://www.cookbook-r.com/Graphs/index.html](http://www.cookbook-r.com/Graphs/index.html)


Similar to the first link, but more detailed. Lots of example plots. 

[http://www.sthda.com/english/wiki/ggplot2-essentials](http://www.sthda.com/english/wiki/ggplot2-essentials)


All of the package functions on one page, with examples

[http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)










