---
title: "Heatmaps in R"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)
```

Load all the required packages.

```{r, message=FALSE}
library(survival)
library(maps)
library(maptools)
library(mapdata)
library(ggplot2)
library(plyr)
library(grid)
library(scatterpie)
```

# Import the datasets into R
```{r}
## Data for mapping:
load('fake-dorado-map-data.RData', verbose=TRUE)
head(fake.dorado)

## EEZ Data
#eez overlay
eez        <- read.table("EZNEW2.txt", sep=" ", header=F, skip=0) #All regions
eez_pna  <- read.csv("pna_eez.csv") #PNA regions

## Import continent line dataset
load(file = 'continent-outlines-only.RData', verbose=TRUE)

all_states <- map_data("world2Hires")
```

# A very simple heatmap
```{r}
subdo <- subset(fake.dorado, fake.dorado$Year == 2015)
# how many rows with each year in my original dataset
table(fake.dorado$Year) 
# how many rows in my filtered dataset
table(subdo$Year)
table(subdo$flag_id)
# subset for JP only
subdo_jp <- subset(subdo, subdo$flag_id == 'JP')
table(subdo_jp$flag_id) 
```
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt))
p
```

Make a heatmap of the yellowfin catch
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=yft_mt))
p
```

Make a heatmap of the bigeye catch
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=bet_mt))
p
```

Make a heatmap with a different colour scheme: 
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'Spectral')
p
```
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'YlOrRd')
p
```

Now let's add a new layer with EEZ outlines
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'Spectral') +
  ## added a polygon layer with the EEZs, only drew EEZ outline (fill=NA)
  geom_polygon(data=eez, aes(eez[,1], eez[,2]), size=2, colour="white", fill=NA) +
  ## narrowed the x and y range of the plot to fit our dataset
  coord_cartesian(xlim=c(190,205), ylim=c(2,18)) +
  theme_bw() # make background white instead of grey
p
```

The next step is to add our continent layer: 
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'Spectral') +
  ## added a polygon layer with the EEZs, only drew EEZ outline (fill=NA)
  geom_polygon(data=eez, aes(eez[,1], eez[,2]), size=2, colour="white", fill=NA) +
  ## narrowed the x and y range of the plot to fit our dataset
  coord_cartesian(xlim=c(180,205), ylim=c(2,26)) +
  theme_bw() + # make background white instead of grey + 
  geom_path(data=countrydf, aes(x=lon, y=lat, group=id))
p
```

Draw this instead with the high resolution polygon dataset: 
```{r}
p <- ggplot() +
  geom_tile(data=subdo_jp, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'Spectral') +
  ## added a polygon layer with the EEZs, only drew EEZ outline (fill=NA)
  geom_polygon(data=eez, aes(eez[,1], eez[,2]), size=2, colour="white", fill=NA) +
  ## narrowed the x and y range of the plot to fit our dataset
  coord_cartesian(xlim=c(180,205), ylim=c(2,26)) +
  theme_bw() + # make background white instead of grey + 
  geom_polygon(data=all_states, aes(x=long, y=lat, group = group),
               colour ="white",  fill="grey10" )
p
```

# Making a map panel 
```{r}
p <- ggplot() +
  geom_tile(data=subdo, aes(x=longitude, y=latitude, fill=alb_mt)) +
  scale_fill_distiller(palette = 'Spectral') +
  ## added a polygon layer with the EEZs, only drew EEZ outline (fill=NA)
  geom_polygon(data=eez, aes(x=V1, y=V2), size=2, 
               colour="white", fill=NA) +
  ## narrowed the x and y range of the plot to fit our dataset
  coord_cartesian(xlim=c(190,205), ylim=c(2,19)) +
 # coord_equal() +
  theme_bw() +
  facet_wrap(~flag_id)
p
```

# Exercises
Doing the same figure but with my Dorado extract.
Make sure your Dorado folder (e.g. MH) in in the ggmaps folder.
Note: for Dorado extracts only, need to specify `fileEncoding='UTF-8-BOM'`

## Import my Dorado data
```{r, eval=FALSE}
dorado.dat <- read.csv('PW/PW custom report 100.csv', fileEncoding = 'UTF-8-BOM')
```

## Filter for 2015
