---
title: "Introduction to the ggplot2 package -- session 2"
output:
  html_document:
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
# Global options
#
# echo = TRUE. Show the code
# comment = "". Don't put anything in front of results from code
# warning = FALSE. Don't display warning messages
# cache = TRUE. Only process new code when you knit
# dev = 'png'. Save graphics in figure folder as png
# dpi = 500. Resolution for saved figures
# fig.height = 3. Height of figures (in inches)

knitr::opts_chunk$set(echo = TRUE, 
                      comment = "",
                      warning = FALSE,
                      cache = TRUE,
                      dev = 'png', 
                      dpi = 500, 
                      fig.height = 3)
```

```{r, message=FALSE}
## Start by loading the data
ce <- read.csv(file ='sample.csv')
head(ce)
## Load required packages
library(ggplot2)
library(tidyr) # package to format dataset
```

# Barplots

Each month there is a certain number of vessels that went fishing. What is the distribution of this number of vessels?

You can find the distribution using the `table` function

```{r table of vessel counts}
table(ce$Vessels)
```

The function `geom_bar` will do this for you and plot up the results. 

```{r plot vessel count}
ggplot(data = ce, aes(x = Vessels)) +
  geom_bar()
```

You'll see the geom_bar function with an `identity` argument. In this case it doesn't sum up the counts first to find the distribution, but uses the user-provided count variable in the data frame. 

A little example will demonstrate this. Note that is this case for ggplot you need to say what will be along the x-axis and y-axis (the counts).

```{r}
df <- data.frame(Year = rep(2013:2017, each=2), 
                 Season=rep(c('Summer','Winter'),5),
                 catch = c(4, 11, 3, 8, 1, 6, 2, 4, 3, 12))
df
ggplot(data = df, aes(x = Year, y = catch)) +
  geom_bar(stat = "identity")
```

```{r}
ggplot(data = df, aes(x = Year, y = catch, fill=Season)) +
  geom_bar(stat = "identity")
```

Show the proportional barplot instead: 
```{r}
ggplot(data = df, aes(x = Year, y = catch, fill=Season)) +
  geom_bar(stat = "identity", position='fill')
```

## Exercises


(1. For albacore tuna find the annual CPUE (as metric tonnes divided by number of hooks). Make a scatter plot of the albacore CPUE over time.)

2. Load the `catch_effort` dataset (from **catch.RData**). Make a line plot of the bigeye catch per year.

Step 1: Import the data and look at it
```{r}
load('catch.RData', verbose=TRUE)
head(catch_effort)
```

Step 2: Set-up `ggplot`, specify which dataset to use with `data = ...`
Next specify the x-variable and the y-variable
Next specify which type of geom layer to use
```{r}
ggplot(data=catch_effort, aes(x=yy, y=BET)) + geom_line()
```

3. For this same dataset, make a barplot of the total catch per year.

Step 1: Find my dataset and format it
DONE.
Step 2: 
```{r}
ggplot(data=catch_effort, aes(x=yy, y=total)) + 
  geom_bar(stat = "identity", fill='azure3')
```

```{r}
ggplot(data=catch_effort, aes(x=yy, y=total)) + 
  geom_bar(stat = "identity", fill='azure3') +
  geom_smooth(method='lm', col='indianred', se = FALSE)
```

4. Now, make a barplot of the total catch per year, but split by tuna species.
Ps: To do this, you will need to first convert the dataset to have tuna species in a single column instead of 3 columns. We use the function `gather` to do this, and create a new column called `Species` and a second column called `catch` with the species catch.

```{r}
## use the function gather in the library tidyr
catchlong <- gather(catch_effort, Species, catch, ALB:YFT)
head(catchlong)
```

Verify that `catchlong` has 3 times the number of rows than `catch_effort` did (since we turned the three tuna columns into a single longer one).

You are now ready to do the stacked barplot. Ps: the colours won't be right, don't worry about this for now.
```{r}
ggplot(data=catchlong, aes(x=yy, y=catch, fill=Species)) + 
  geom_bar(stat = "identity")
```


5. Do this same plot but as a proportional barplot.
```{r}
ggplot(data=catchlong, aes(x=yy, y=catch, fill=Species)) + 
  geom_bar(stat = "identity", position='fill')
```

6. Using the same data, plot albacore, bigeye and yellowfin catch as lines on the same plot

```{r}
ggplot(data=catchlong, aes(x=yy, y=catch, color=Species)) + 
  geom_line()
```

# Plots broken up by groups
One of the great features of `ggplot` is that it is very easy to create panel plots, using the functions `facet_wrap` and `facet_grid`.

## Panel plots by one variable

You can make separate plots of the albacore tuna catch density by month, by adding in the `facet_wrap` function.

You put `Month` in the facet_wrap function, preceding it by a tilde ("~").

```{r plot density by month}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_density(adjust = 0.5, fill = "darkblue") +
  facet_wrap( ~Month)

```

ps: you can only split plots by a discrete (categorical) variable, e.g. Month, season, species, etc.

Plot the distribution of albacore, split by year
```{r plot density by month II}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_density(adjust = 0.5, fill = "darkblue") +
  facet_wrap( ~Yr)
```

Do this same plot but with a histogram:
```{r}
ggplot(data = ce, aes(x = ALB.mt)) +
  geom_histogram() +
  facet_wrap( ~Yr)
```

Make a line plot of albacore catch per month, split by year panels.
```{r}
ggplot(data = ce, aes(x = Month, y=ALB.mt)) +
  geom_line() +
  facet_wrap( ~Yr)
```

Make a line plot of bigeye catch per month, split by year panels.
```{r}
ggplot(data = ce, aes(x = Month, y=BET.mt)) +
  geom_line() +
  facet_wrap( ~Yr)
```

## Panel plots by two variables

If there are two variables by which you would like to visualise your data, use the function `facet_grid` instead, specifying the two variables to split the panels on each side of the tilde.

First we will add a second categorical variable to split out dataset.
```{r}
ce$Season <- ifelse(ce$Month %in% c(1:3,10:12), 'Cool', 'Warm')

head(ce)
```

Now we make a plot of albacore catch distribution by year and season:
```{r, fig.height=10}
ggplot(data = ce, aes(x = ALB.mt, fill=Season)) +
  geom_density(adjust = 0.5) +
  facet_grid(Yr~Season)
```

## Exercises

1. Make the density plot of bigeye catch, but split in panels by Month.
```{r}
ggplot(data = ce, aes(x = BET.mt)) +
  geom_density(adjust = 0.5, fill = "darkblue") +
  facet_wrap( ~Month)
```

2. Make the histogram of albacore CPUE, but split in panels by season.
```{r}
# add new albacore CPUE column to the ce dataframe 
# CPUE = catch / effort
ce$ALB.cpue <- ce$ALB.mt/ce$Hooks.100 # add new column in dataset with the ALB CPUE
ggplot(ce, aes(x=ALB.cpue)) +
  geom_histogram() +
  facet_wrap(~Season)
```

3. Make a scatterplot of bigeye vs. yellowfin catch, but split in panels by season, with the years coloured in individually.
```{r, fig.width=8}
ggplot(data=ce, aes(x=BET.mt, y=YFT.mt, color=as.factor(Yr))) + 
  geom_point() +
  facet_wrap(~Season)
```


# Juxtaposing lines on the same plot

Read in new format of dataset with species in the column 'Species' instead of one column per species type, add date to plot
```{r}
ce.long <- read.csv('sample_data-long.csv')
names(ce.long)
table(ce.long$Species)
ce.long$date <- ce.long$Yr + (ce.long$Month-1)/12
head(ce.long)
```
Plot catch (in mt) by date using ggplot. The group argument allows to separate line by the values in an index column (here **Species**).
```{r}
ggplot(ce.long, 
       aes(x=date, y=Catch.mt, color=Species)) + 
  geom_line() + 
  geom_point() 
```

To change the color scheme, you can use the function `scale_color_brewer()`.
```{r}
ggplot(ce.long, 
       aes(x=date, y=Catch.mt, color=Species)) + 
  geom_line() + 
  geom_point() + 
  scale_color_brewer(palette='Spectral')
```

Plot catch (in mt) for each species in a single panel.
```{r}
ggplot(ce.long, 
       aes(x=date, y=Catch.mt)) + 
  geom_line(colour='dodgerblue4') + 
  geom_point() + 
  scale_color_brewer(palette = 'Spectral') + 
  theme_bw() + facet_wrap(~Species, scales ='free_y')
```

Plot catch (in mt) for each species in a single panel and colour the points as a function of the effort (`Hooks.100`) in that month.
```{r}
ggplot(ce.long, 
       aes(x=date, y=Catch.mt, colour=Hooks.100)) + 
  geom_line(colour='red') + 
  geom_point() + 
  theme_bw() + facet_wrap(~Species, scales ='free_y')
```


